Geospatial - Assignment 3

### 2025-03-02 hw3/hw3_MoritzPeistNoemiLucchiSimonVellin.Rmd
Refined 2nd part

---
title: "Geospatial - Assignment 3"
author:
- Moritz Peist
- Noemi Lucchi
- Simon Vellin
output:
  html_document:
    toc: true
    df_print: paged
  pdf_document:
    toc: true
  html_notebook:
    toc: true
---

@@

# Disable warning messages and SVG plots
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  # fig.ext = "svg",
  fig.width = 16,
  fig.height = 8,
  fig.retina = 2,
  dpi = 300
  # dev = "svg"
)
```

@@

To calculate the distances between the 10 top populated cities in Spain, and represent the roads within the continental Spain, we used data from the [Natural Earth website](https://www.naturalearthdata.com/downloads/10m-cultural-vectors/). Precisely, we loaded:
  - ne_10m_populated_places shapefile for the most populated cities 
  - ne_10m_roads for the streets

@@

From the R package spData, we use the world dataset to extract the geometry of Spain, filtering only for continental Spain. Then, from the dataset containing the most populated cities, we extract only the cities in Spain, sort them in decreasing order of population, and retain only the top 10. Finally, from these top 10 cities, we extract Madrid and Vigo, which will be used in the plot.

@@

```{r}
# Plotting
ggplot() +
  geom_sf(data = st_transform(spain, 4326), fill = "lightgray") +
  geom_sf(data = st_transform(spain_roads, 4326), color = "black", size = 0.3) +
  geom_sf(data = st_transform(spain_top_10_cities, 4326), size = 1, color = "black") +
  geom_sf(
    data = st_transform(madrid_vigo %>% filter(NAME == "Madrid"), 4326),
    aes(shape = "Madrid", color = "Madrid"), size = 3
  ) +
  geom_sf(
    data = st_transform(madrid_vigo %>% filter(NAME == "Vigo"), 4326),
    aes(shape = "Vigo", color = "Vigo"), size = 3
  ) +
  scale_color_manual(
    name = "Cities",
    values = c("Madrid" = "red", "Vigo" = "turquoise")
  ) +
  scale_shape_manual(
    name = "Cities",
    values = c("Madrid" = 16, "Vigo" = 17)
  )
```

@@

Using the distance matrix calculated in Part A, we now assess the isolation of Madrid and Vigo by analyzing their bilateral distances to the other top 10 most populated cities in Spain. We begin by extracting the distances for Madrid and Vigo.

```{r}
# Extract city names from the top 10 cities (excluding Madrid and Vigo for isolation analysis)
other_cities <- spain_top_10_cities$NAME[!spain_top_10_cities$NAME %in% c("Madrid", "Vigo")]

# Extract distances from Madrid to other cities
madrid_distances <- dist_matrix["Madrid", other_cities]

# Extract distances from Vigo to other cities
vigo_distances <- dist_matrix["Vigo", other_cities]
```

@@

```{r}
# Create a data frame with distances for plotting
density_data <- data.frame(
  City = c(rep("Madrid", length(madrid_distances)), rep("Vigo", length(vigo_distances))),
  Distance = c(madrid_distances, vigo_distances)
)

# Plot smoothed density distributions using geom_density()
density_plot <- ggplot(density_data, aes(x = Distance, color = City, fill = City)) +
  geom_density(alpha = 0.3) +  # Add transparency for overlap
  scale_color_manual(values = c("Madrid" = "red", "Vigo" = "turquoise")) +
  scale_fill_manual(values = c("Madrid" = "red", "Vigo" = "turquoise")) +
  labs(
    title = "Smoothed Density Distributions of Distances",
    x = "Distance (km)",
    y = "Density",
    color = "City",
    fill = "City"
  ) +
  theme_minimal()

@@

```{r}
# Extract distances for Madrid and Vigo to all cities, including themselves (self-distance = 0)
madrid_distances_all <- dist_matrix["Madrid", ]
vigo_distances_all <- dist_matrix["Vigo", ]

# Create a data frame with distances for plotting, including self-distances
density_data_all <- data.frame(
  City = c(rep("Madrid", length(madrid_distances_all)), rep("Vigo", length(vigo_distances_all))),
  Distance = c(madrid_distances_all, vigo_distances_all)
)

# Plot smoothed density distributions using geom_density(), including self-distances
density_plot_all <- ggplot(density_data_all, aes(x = Distance, color = City, fill = City)) +
  geom_density(alpha = 0.3) +  # Add transparency for overlap
  scale_color_manual(values = c("Madrid" = "red", "Vigo" = "turquoise"), labels = c("Madrid", "Vigo")) +
  scale_fill_manual(values = c("Madrid" = "red", "Vigo" = "turquoise"), labels = c("Madrid", "Vigo")) +
  labs(
    x = "distance",
    y = "density",
    color = "origin",
    fill = "origin"
  ) +
  theme_minimal() +
  theme(
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  )

>>>>

---
title: "Geospatial - Assignment 3"
author:
  - Moritz Peist
  - Noemi Lucchi
  - Simon Vellin
output:
  html_document:
    toc: true
    df_print: paged
  html_notebook:
    toc: true
  pdf_document:
    toc: true
---

@@

# Disable warning messages and SVG plots
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  # fig.ext = "svg",
  fig.width = 16,
  fig.height = 8,
  fig.retina = 2,
  dpi = 300
  #dev = "svg"
)
```

@@

To calculate the distances between the 10 top populated cities in Spain, and represent the roads within continental Spain. We used data from the [Natural Earth website](https://www.naturalearthdata.com/downloads/10m-cultural-vectors/). Precisely, we loaded:

- ne_10m_populated_places shapefile for the most populated cities 
- ne_10m_roads for the streets

@@

From the R package spData, we use the world dataset to extract the geometry of Spain, filtering only for continental Spain i.e., we exclude roads outside continental Spain which seems to not have happened in the initial graph were roads overflow frontiers. Then, from the dataset containing the most populated cities, we extract only the cities in Spain, sort them in decreasing order of population, and retain only the top 10. Finally, from these top 10 cities, we extract Madrid and Vigo, which will be used in the plot.

@@

```{r}
# Plotting
ggplot() +
  geom_sf(data = st_transform(spain, 4326), fill = "lightgray") +
  geom_sf(data = st_transform(spain_roads, 4326), color = "black", size = 0.3) +
  geom_sf(data = st_transform(spain_top_10_cities, 4326), size = 1, color = "black") +
  geom_sf(
    data = st_transform(madrid_vigo %>% filter(NAME == "Madrid"), 4326),
    aes(shape = "Madrid", color = "Madrid"), size = 3
  ) +
  geom_sf(
    data = st_transform(madrid_vigo %>% filter(NAME == "Vigo"), 4326),
    aes(shape = "Vigo", color = "Vigo"), size = 3
  ) +
  scale_color_manual(
    name = "Cities",
    values = c("Madrid" = "red", "Vigo" = "blue")
  ) +
  scale_shape_manual(
    name = "Cities",
    values = c("Madrid" = 16, "Vigo" = 17)
  )
```

@@

Using the distance matrix calculated in Part A, we now assess the isolation of Madrid and Vigo by analyzing their bilateral distances to the other top 10 most populated cities in Spain. We begin by extracting the distances for Madrid and Vigo. Note, however, that for the average distance calculation, that we exclude the cities from the calculations with themselves to exclude the self-distance of 0.

```{r}
# Extract city names from the top 10 cities (excluding Madrid and Vigo for isolation analysis)
df_dist <- as.data.frame(dist_matrix) %>%
  rownames_to_column(var = "origin") %>%
  # Set up long format of the matrix
  gather(key = "destination", value = "distance", -origin) %>%
  # Exclude all other cities
  filter(origin %in% c("Madrid", "Vigo")) %>% 
  # Order first by origin, then by destination
  arrange(origin, destination)

# Now exclude the self-distances
df_dist_wo_self <- df_dist %>%
  filter(origin != destination)

# Extract distances from Madrid to other cities
madrid_distances <- df_dist_wo_self %>%
  pull(distance)

# Extract distances from Vigo to other cities
vigo_distances <- df_dist_wo_self %>%
  pull(distance)
```

@@

```{r}
# Plot smoothed density distributions using geom_density()
density_plot <- ggplot(df_dist_wo_self, aes(x = distance, color = origin, fill = origin)) +
  geom_density(alpha = 0.3) +  # Add transparency for overlap
  scale_color_manual(values = c("Madrid" = "red", "Vigo" = "turquoise")) +
  scale_fill_manual(values = c("Madrid" = "red", "Vigo" = "turquoise")) +
  labs(
    title = "Smoothed Density Distributions of Distances",
    x = "Distance (km)",
    y = "Density",
    color = "City",
    fill = "City"
  ) +
  theme_minimal()

@@

```{r}
# Plot smoothed density distributions using geom_density(), including self-distances
density_plot_all <- ggplot(df_dist, aes(x = distance, color = origin, fill = origin)) +
  geom_density(alpha = 0.3) +  # Add transparency for overlap
  scale_color_manual(values = c("Madrid" = "red", "Vigo" = "turquoise"), labels = c("Madrid", "Vigo")) +
  scale_fill_manual(values = c("Madrid" = "red", "Vigo" = "turquoise"), labels = c("Madrid", "Vigo")) +
  labs(
    x = "distance",
    y = "density",
    color = "origin",
    fill = "origin"
  ) +
  theme_minimal() +
  theme(
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  )
